Allee Effect and Cultural Evolution

### 2015-03-24 manuscript/draft.tex
added references

Formal models portraying changes in human subsistence have been mostly offered in human behavioral ecology (Smith and Winterhalder 1992, Bird and O'Connell 2006, Kennet and Winterhalder 2006), where several mathematical models inspired from evolutionary ecology offered precise prediction based on expected patterns under the assumption of optimality (Belovsky 1988)\footnote{citation needed}. While earliest models relied almost exclusively on this assumption, most recent works are increasing discussing how the knowledge of alternative subsistence strategies are acquired, and whether this process of learning can drive the evolutionary trajectory of a given population (Henrich and Boyd, 1998), partly as a synergy with the dual inheritance theory (Boyd and Richerson 1985).

The results of these works often exhibits counter-intuitive outcomes that cannot be predicted by standard assumptions of optimality. For instance, extreme reliance on social learning can promote information parasitism (Giraldeau et al 2002), leading to a decline in the proportion of asocial learners that in certain environmental context can have strongly detrimental effects at the population level (Whitehead and Richerson 2009). Other examples include instances where a reduced effective population size promotes an increase in random drift, and the ultimate loss of adaptive traits (Henrich 2004), or instances where neutral traits are jointly transmitted together with adaptive ones (Ackland et al 2007). These examples illustrate how theoretical models based on the dual inheritance theory can extend predictions of the standard optimal foraging theory, integrating dynamics that, at least at a first glance,  are regarded as counter-intuitive. These include episodes of reversions (e.g. \footnote{citation needed}), where a more recent optimal strategies are abandoned for older suboptimal variants; stable coexistence (e.g. \footnote{citation needed}), where optimal and suboptimal strategies co-exist within a population; and even the complete loss of optimal variants (e.g. Henrich 2004). The evolutionary processes leading to these different outcomes are undoubtedly contingent to local historical dynamics, nonetheless we argue that a simple model of social learning in combination with a positive and negative frequency dependent selection can lead all these equilibria.

@@

A key feature that underlies many models of cultural transmission is that the social learner is often guided by some form of evaluation of characteristics associated with potential teachers or demonstrators. To put it simply, one can decide to copy traits possessed by individuals showing some signal that suggests higher payoff, success, or fitness (e.g. number of offspring, income, presence/absence of key traits, etc.). This learning strategy has been broadly referred to as directed social learning (Coussi-Korbel and Fragaszy 1995), model-biased transmission (Boyd and Richerson 1985, Henrich and Gil-White 2001), attraction model (Acerbi and Bentley 2014), and payoff-based transmission\footnote{citation needed}. The exact details of these model differ to some extent, as some might imply that the target trait of interest and the observed payoff-signal is only weakly correlated (e.g. prestige biased transmission), while other entails a more direct link between the two. In one sense one can argue that they are part of a broader category of social learning strategy (see also Crema and Lake, under review).

The degree of correlation between the payoff-signal and the target trait can also be viewed as a form of uncertainty or amount of information content offered by the signal. Previous studies (Schlag 1998,1999; Baldini 2013, Lake and Crema 2012) have explored how different learning strategies (e.g. copy-the-best, or copy-the-best-average, or copy-proportionally) show different dynamics, and are differently biased by this uncertainty. In some cases these might produce counter-intuitive outcomes. For instance, Baldini (2013) has shown that when the learning strategy is  copy-the-best suboptimal variants with higher payoff uncertainty and lower mean (equivalent to a lower correlation between signal payoff-signal and the target trait) can be favored over variants with higher mean and lower uncertainty. In other contexts, the uncertainty might be in the identification of the correct target trait, which could lead to the adoption of variants that have minimal or no correlation with the observed cue (cf. “evolutionary hitchhiking”,  Ackland et al 2007).

@@

Several cultural traits are expected to show such density dependence. For example, behavior linked to the exploitation or consumption of limited resources are expected to exhibit a negative frequency dependence.  Novel optimal variants are likely to spread rapidly when the trait is rare, but once a threshold frequency is exceeded,  beneficial effect will start to decline,  the payoff will become smaller,  and eventually rarer suboptimal variants can be preferred (see Lake and Crema 2012 for an extensive analysis).  Negative dependency is however not the only form frequency dependence we should expect. In some cases, cultural variants that are strongly based on direct or indirect cooperation might require a minimum frequency to provide sufficient fitness, or more broadly show a positive correlation with the number of individuals possessing the trait. This positive frequency dependence, known in ecology as Allee Effect \citep{allee1958}, can be expected at different meta-population levels.  For example foraging groups are known to show higher performance when the number of individuals is not too small (Smith 1981), while certain activity such as anthropogenic fire (Bird et al 2013) or selective hunting (Dods 2002) might require a sufficient population density to promote sufficient niche enhancement and a consequent increase in payoff (Rowley-Conwy 2011). More\footnote{citation needed!} Language? Bettencourt Cities? \enrico{Ideally we need more example, not necessarily within the domain of subsistence activities. Any Allee Effect? Add Ostrom 2000 mention collective good problems here}.

@@

\enrico{We need to mention that some behaviour might actually enhance the other community (or be detrimental). This can be modelled using a competition coefficient as in Jang 2013. I don't think we need to explore this but it is definitely worth mentioning!}

>>>>

Formal models portraying changes in human subsistence have been mostly offered in human behavioral ecology \citep{smith1992,bird2006,kennett2006}, where several mathematical models inspired from evolutionary ecology offered precise prediction based on expected patterns under the assumption of optimality \citep{belovsky1988}. While earliest models relied almost exclusively on this assumption, most recent works are increasing discussing how the knowledge of alternative subsistence strategies are acquired, and whether this process of learning can drive the evolutionary trajectory of a given population \citep{henrich1998}, partly as a synergy with the dual inheritance theory \citep{boyd1985}.

The results of these works often exhibits counter-intuitive outcomes that cannot be predicted by standard assumptions of optimality. For instance, extreme reliance on social learning can promote information parasitism \citep{giraldeau2002}, leading to a decline in the proportion of asocial learners that in certain environmental context can have strongly detrimental effects at the population level \citep{whitehead2009}. Other examples include instances where a reduced effective population size promotes an increase in random drift, and the ultimate loss of adaptive traits \citep{henrich2004}, or instances where neutral traits are jointly transmitted together with adaptive ones \citep{ackland2007}. These examples illustrate how theoretical models based on the dual inheritance theory can extend predictions of the standard optimal foraging theory, integrating dynamics that, at least at a first glance, are regarded as counter-intuitive. These include episodes of reversions (e.g. \footnote{citation needed}), where a more recent optimal strategies are abandoned for older suboptimal variants; stable coexistence (e.g. \footnote{citation needed}), where optimal and suboptimal strategies co-exist within a population; and even the complete loss of optimal variants \citep[e.g.][]{henrich2004}. The evolutionary processes leading to these different outcomes are undoubtedly contingent to local historical dynamics, nonetheless we argue that a simple model of social learning in combination with a positive and negative frequency dependent selection can lead all these equilibria.

@@

A key feature that underlies many models of cultural transmission is that the social learner is often guided by some form of evaluation of characteristics associated with potential teachers or demonstrators. To put it simply, one can decide to copy traits possessed by individuals showing some signal that suggests higher payoff, success, or fitness (e.g. number of offspring, income, presence/absence of key traits, etc.). This learning strategy has been broadly referred to as directed social learning \citep{coussi1995}, model-biased transmission \citep{boyd1985,henrich2001}, attraction model \citep{acerbi2014}, and payoff-based transmission\footnote{citation needed}. The exact details of these model differ to some extent, as some might imply that the target trait of interest and the observed payoff-signal is only weakly correlated (e.g. prestige biased transmission), while other entails a more direct link between the two. In one sense one can argue that they are part of a broader category of social learning strategy\footnote{citation needed - see also Crema and Lake, under review}.

The degree of correlation between the payoff-signal and the target trait can also be viewed as a form of uncertainty or amount of information content offered by the signal. Previous studies \citep{schlag1998,baldini2013,lake2012} have explored how different learning strategies (e.g. copy-the-best, or copy-the-best-average, or copy-proportionally) show different dynamics, and are differently biased by this uncertainty. In some cases these might produce counter-intuitive outcomes. For instance, \citet{baldini2013} has shown that when the learning strategy is  copy-the-best suboptimal variants with higher payoff uncertainty and lower mean (equivalent to a lower correlation between signal payoff-signal and the target trait) can be favored over variants with higher mean and lower uncertainty. In other contexts, the uncertainty might be in the identification of the correct target trait, which could lead to the adoption of variants that have minimal or no correlation with the observed cue \citep[cf. "evolutionary hitchhiking"][]{ackland2007}.

@@

Several cultural traits are expected to show such density dependence. For example, behavior linked to the exploitation or consumption of limited resources are expected to exhibit a negative frequency dependence. Novel optimal variants are likely to spread rapidly when the trait is rare, but once a threshold frequency is exceeded,  beneficial effect will start to decline, the payoff will become smaller, and eventually rarer suboptimal variants can be preferred (see Lake and Crema 2012 for an extensive analysis). Negative dependency is however not the only form frequency dependence we should expect. In some cases, cultural variants that are strongly based on direct or indirect cooperation might require a minimum frequency to provide sufficient fitness, or more broadly show a positive correlation with the number of individuals possessing the trait. This positive frequency dependence, known in ecology as Allee Effect \citep{allee1958}, can be expected at different meta-population levels. For example foraging groups are known to show higher performance when the number of individuals is not too small\footnote{There was a reference on Smith 1981...is this JM Smith 'evolution and theory of games'? This is 1982}, while certain activity such as anthropogenic fire \citep{bird2013} or selective hunting (Dods 2002) might require a sufficient population density to promote sufficient niche enhancement and a consequent increase in payoff \citep{rowley2011}. More\footnote{citation needed!} Language? Bettencourt Cities? \enrico{Ideally we need more example, not necessarily within the domain of subsistence activities. Any Allee Effect? Add Ostrom 2000 mention collective good problems here\footnote{what is ostrom 2000?}}.

@@

\enrico{We need to mention that some behaviour might actually enhance the other community (or be detrimental). This can be modelled using a competition coefficient as in \citep{jang2013}. I don't think we need to explore this but it is definitely worth mentioning!}
